Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-07 talk/pepm2011/paper.tex
add which languages pypy supports

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C. XXX languages that are done using PyPy

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer by a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter, and
the process benefits all languages implemented as an interpreter in RPython.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning.

This trace of operations is then the basis of the generated code. The trace is
optimized in some ways, and then turned into machine code. Generating machine
code is simple, because the traces are linear and the operations are very close
to machine level. The trace corresponds to one concrete execution of a loop,
therefore the code generated from it is only one possible path through the loop.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \texttt{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \texttt{if} statement is true, because if
it isn't, the rest of the trace is not valid.

@@

The creation of objects in category 1 is removed by the optimization described
in Section~\ref{sec:statics}. XXX

>>>>

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an \emph{interpreter}
for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way that type inference becomes
possible. The language interpreter can then be compiled (``translated'') with
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

A number of languages have been implemented with PyPy. The project was started
to get a better Python implementation, which inspired the name of the project
and is still the main focus of development. In addition a number of other
languages were implemented, among them a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010}, a Smalltalk VM
\cite{carl_friedrich_bolz_back_2008} and a GameBoy emulator
\cite{bruni_pygirl:_2009}.

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is mostly
automatic; it only needs to be guided by the language implementer by a small number of
source-code hints. Mostly-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process.
By construction, the generated JIT has the same semantics as the interpreter.
Many optimizations can benefit all languages implemented as an interpreter in RPython.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

This trace of operations is then the basis of the generated code. The trace is
optimized in some ways, and then turned into machine code. Both optimizations
and machine code generation is simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through it.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \texttt{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \texttt{if} statement is true, because if
it isn't, the rest of the trace is not valid.

@@

The creation of objects in category 1 is removed by the optimization described
in Sections~\ref{sec:statics} and \ref{sec:formal}. We will look at objects in
category 3 in Section~\ref{sec:crossloop}.
